Spontaneous Similarity Judgments of Musical Tones played with various Playing Techniques

### 2017-12-12 paper/lagrangeSpontaneousSimilarity2017.tex
work in progress

Musical timbre is an interesting notion to study as it is perceptively defined. Thus, by studying musical timbre, one study many facets of the human auditory system which responds to a diverse -- but somewhat more controled than other stimuli such as speech or environemntal sounds -- set of physical stimuli.

A popular view of the kind of stimuli we consider is that they can be described by a spread of energy that evolve through time and across frequency. By following this trend, most researchers translate the negation of the ASA definition of timbre into a more relative definition : musical timbre is about the distribution of intensity across the time / frequency plane.

A great deal of litterature focuses on the instaneous spectrum, \textit{i.e.} the timbre can be defined without considering evolution of the energy distribution trough time. Other studies demonstrated the importance of the onset in the recognition of some musical instruments. Those studies on a rather short term model where the waveform is observed in a range that do not exceed 100 ms.

Interestingly, modern physiological models of the primary mamalian auditory system also considers the evolution of the energy distribution at longer time scales, \textit{i.e.} how the energy modulates through time for given frequencies.  Dau's model acount for frequency and rate of modulation. Shamma's model account for frequency, rate and bandwidth to fully describes the physiological evidence gathered by studying the auditory system of the ferret. Though, most of the numerical experiments done using this model did not fully demonstrated the usefullness of this third dimension.

Thus, by taking a signal processing view of the matter, one could consider that a first stage of the auditory system proceeds to a rather fixed set of convolutions across time and frequency, projecting the signal into a large set of descriptive features organized across frequency and rate of modulations.

What the higher level of the auditory cortex does remains largely unkown, but we can assume that the level of plasticity is very high, as any living entity has to adapt their percepts to a wide diversity of tasks.  Considering the lack of information about this stage, we will assume in this paper that this second step is rather opportunistic, in the sense that, from this large set of features, it aims at multiplexing in some way the ones that are the most relevant for the task at hand.

Considering not only the type of musical instrument but also the playing technique is interesting in that matter, as it invite us to consider the relation between the rate of modulation and timbre percepts. A notion that have not been extensively studied in the litterature. Compared to instrument class, the reference that is usually taken in timbre studies, it also permits to consider the notion of timbral similarity at finer grain of detail.

@@

The dataset considered in this study has audio recording of 16 different  musical instruments () played with different playing techniques, which leads to 143 different couple of instrument / playing technique. For each couple, the pitch and intonation is varied leading to 25444 samples. The way the dataset is segmented is explained in more details in Appendix \ref{sec:dataset}.

@@

There is a clear gain of NMI when considering the playing technique which means that the subjects primarily used cues corresponding to the playing technique to organize the \ipt. Also, for both curves, a global trend of increase NMI with respect to the number of clusters in the $j$ partition can be observed. This is a well documented fact, which simply put, means that it is easier to match a given partition with more clusters at hand. To compensate for this bias, one can consider the NMI that would be achieved by comparing several random partitions of increasing number of clusters to the reference one, be it the instrument partition ($nullI$) or the playing technique partition ($nullPt$). The resulting averaged NMI for 100 randomly generated partitions are displayed on Figure \ref{fig:clusters}. Those curves are useful for normalizing $j/I$ and $j/Pt$ by simply substracting the so)called null NMIs:
\begin{equnarray}
  nj/I &=& j/I - nullI  \\
  nj/Pt &=& j/Pt - nullPt  \\
\end{equnarray}
This normalization is useful for identifying more easily the number of clusters for which the comparison is most relevant. As can be seen on Figure \ref{fig:clusters}, the number of clusters that leads to the highest NMI is 5 for the playing technique reference and 8 for the instrument reference, leading respectively to an NMI of $0.46$ and $0.21$. For the latter, a smaller number down to 6 do not decrease the NMI by much, a fact that will be latter discussed in more details. Even by considering the null normalization, the matching level is much higher for the playing technique refereence.

@@

\section{Dataset}

distribution of instruments and modes

\subsection{Naming}

abbreviations 4 instruments and modes.

>>>>

Musical timbre is an interesting notion to study as it is perceptively defined \cite{grey1977multidimensional}. Thus, by studying musical timbre, one study many facets of the human auditory system which responds to a diverse set of physical stimuli. However this diversity can be more easily more controlled than other stimuli such as speech or environmental sounds.

The focus of this paper is to study the influence of the playing technique on the perception of timbre.

A popular view of the kind of stimuli we consider is that they can be described by a spread of energy that evolve through time and across frequency. By following this trend, most researchers translate the exclusion definition of timbre \cite{marozeau2003dependency} into a more specified definition. The exclusion definition states that timbre is "the fourth component of sound quality, the first three being pitch, loudness and duration". In an attempt to specify more this notion of timbre, we will loosely consider in this paper that musical timbre is the influence on the human auditory system of the distribution of intensity across the time / frequency plane.

A great deal of litterature focuses on the instantaneous spectrum \cite{grey1978perceptual}. Other studies also demonstrated the importance of the onset in the recognition of some musical instruments. All those studies generally assume a rather short term model where the waveform is observed in a range that do not exceed 100 ms, a common porcessing scenario that is very commonly used in computational recognition models \cite{tzanetakis2002musical}.

Agus \& al effectively showed that a great deal of instrument recognition can be achieved with very little listening time \cite{agus2012fast}. Though, instrument recognition is only part of musical timbre perception. We propose in this paper to study the influence of the playing technique on musical timbre percept. Evidently, the influence of the playing technique can be measured in terms of energy distribution along the frequency axis but also with respect to time on long time scales.

Interestingly, modern physiological models of the primary mamalian auditory system also considers the evolution of the energy distribution at longer time scales, \textit{i.e.} how the energy modulates through time for given frequencies.  Dau's model acounts for  rate of modulation over frequency bands of fixed bandwidth on a logarithmic frequency scale\cite{dau1997modeling}. Shamma's model accounts for rate  of modulations over several bandwidth to fully describes the physiological evidence gathered by studying the auditory system of the ferret \cite{yang1992auditory}. Though, some numerical applications of this model did not demonstrated the usefullness of the bandwidth dimension \cite{mesgarani2006discrimination}.

Thus, by taking a signal processing view of the matter, one could consider that a first stage of the auditory system proceeds to a rather fixed set of convolutions across time and frequency, projecting the signal into a large set of descriptive features organized across frequency and rate of modulations \cite{anden2014deep}.

What the higher levels of the auditory cortex does remains largely unkown, but we can assume that the degree of freedom of the underlying model is very large contrary to the early stages.  We will assume in this paper that this second step is rather opportunistic. From the large set of features it recieves, this second stage aims at multiplexing in some way the ones that are the most relevant for the task at hand. Related work includes the use of the auditory model proposed by Prof. Shamma for studying musical timbre \cite{patil2012music}.

Considering not only the type of musical instrument but also the playing technique is interesting in that matter, as it invite us to consider the relation between the rate of modulation and timbre percepts. A notion that have not been extensively studied in the litterature \cite{burred2010dynamic}. Compared the reference that is usually taken in timbre studies \textit{i.e.} the instrument class, considering the playing technique allows us to consider the notion of timbral similarity at finer grain of detail.

@@

The dataset considered in this study has audio recording of 16 different musical instruments played with different playing techniques, which leads to 143 different couple of instrument / playing technique. For each couple, the pitch and intonation is varied leading to 25444 samples. The way the dataset is segmented is explained in more details in Appendix \ref{sec:dataset}.

@@

There is a clear gain of NMI when considering the playing technique which means that the subjects primarily used cues corresponding to the playing technique to organize the \ipt. Also, for both curves, a global trend of increase NMI with respect to the number of clusters in the $j$ partition can be observed. This is a well documented fact, which simply put, means that it is easier to match a given partition with more clusters at hand. To compensate for this bias, one can consider the NMI that would be achieved by comparing several random partitions of increasing number of clusters to the reference one, be it the instrument partition ($nullI$) or the playing technique partition ($nullPt$). The resulting averaged NMI for 100 randomly generated partitions are displayed on Figure \ref{fig:clusters}. Those curves are useful for normalizing $j/I$ and $j/Pt$ by simply substracting the so)called null NMIs:
\begin{eqnarray}
  nj/I &=& j/I - nullI  \\
  nj/Pt &=& j/Pt - nullPt  \\
\end{eqnarray}
This normalization is useful for identifying more easily the number of clusters for which the comparison is most relevant. As can be seen on Figure \ref{fig:clusters}, the number of clusters that leads to the highest NMI is 5 for the playing technique reference and 8 for the instrument reference, leading respectively to an NMI of $0.46$ and $0.21$. For the latter, a smaller number down to 6 do not decrease the NMI by much, a fact that will be latter discussed in more details. Even by considering the null normalization, the matching level is much higher for the playing technique refereence.

@@

\section{Dataset} \label{sec:dataset}

The dataset considered in this paper as been recorded at Ircam and is composed of audio recording of 16 different musical instruments played with different playing techniques, which leads to 143 different couple of instrument / playing technique. For each couple, the pitch and intonation is varied leading to 25444 audio samples.

The recorded instruments belongs to the wind and string classes. For winds, the instruments are: accordion (Acc), alto, tuba (Tb), bassoon, clarinet (Cl), flute (Ft), horn (Hn), oboe (Ob), trombone, and trumpet (Tp). For the string class, the instruments are: guitar (Gt), harp, viola (Va), violin (Vn), violoncello (Vc), and contrabass (Cb).

Those instruments are recorded for different nuance and pitch if the latter is relevant, but more importantly, several playing technique are considered.

\section{Dendrogram}

\begin{figure}
\center
\includegraphics[width = \textwidth]{figures/allLinkage.png}
\caption{Dendrogram achiebed using the agglomerative clustering algorithm.}
\label{fig:dendrogram}
\end{figure}
